\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-10 submission/manuscript.tex
update FigS4 and add more to manuscript to send Sarah

Machine learning (ML) modeling of the human microbiome has the potential
to identify the microbial biomarkers and aid in diagnosis of many
chronic diseases such as inflammatory bowel disease, diabetes and
colorectal cancer. Progress has been made towards developing ML models
that predict health outcomes from bacterial abundances, but rigourous ML
models are scarce due to the flawed modeling methods that call the
validity of developed ML models into question. Furthermore, the use of
black box ML models has hindered the validation of microbial biomarkers.
To overcome these challenges, we benchmarked seven different ML models
that use fecal 16S rRNA sequences to predict the presence/absence of
colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
cases). To show the effect of model selection, we assessed the
predictive performance, interpretability, and computational efficiency
of the following models: L2-regularized logistic regression, L1 and L2
support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest, and extreme gradient boosting
(XGBoost). The random forest model was best at detecting CRC lesions
with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to
interpret. Despite its simplicity, L2-regularized logistic regression
followed random forest in predictive performance with an AUROC of 0.680,
and it trained much faster (12 min). This study showed that we should
choose ML models based on our expectations of predictive performance,
interpretability and our computational resources. It also established
standards for modeling pipelines of microbiome-associated ML models.

@@

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. ML models learn
from existing data to predict the outcomes of new data and allow us to
infer on the reasons underlying that prediction. Therefore, researchers
have started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented (12, 13). More notably, we commonly see flawed ML practices
such as using ML pipelines where there is no seperate held-out test
dataset to evaluate model performance or reporting few or only the best
outcomes of cross-validation. Even when there are seperate testing sets
to evaluate model performance, there are large differences between
cross-validation and testing performances that indicate overfitting as
well as large confidence intervals for testing performances (4, 14--20).

Moreover, there is a lack of discussion on why a particular ML model is
utilized. Recently, there is a trend towards using more complex ML
models such as random forest, extreme gradient boosting and neural
networks without a discussion on if and how much model interpretibility
is necessary for the study (11, 21--23). Black box machine learning
models require posthoc explanations to determine the feature importances
in making a prediction. These explanations can be misleading and at
times unreliable when making high-stake decisions about someone's health
(24). The lack of transparency on model selection and interpretation as
well as flawed modeling methods negatively impact model validity and
reproducibiliy. We need to strive toward better machine learning
practices by (1) implementing rigourous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of model accuracy, complexity, interpretibility
and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences of 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development and
fecal 16S rRNA gene sequences have been used to detect CRC. We built
seven ML models using fecal 16S rRNA gene sequences to predict healthy
patients versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We explained the feature importances in non-linear models, SVM with
radial basis kernel, decision tree, random forest and XGBoost, using
permutation importance on the held-out test data where we randomly
permuted non-correlated features individually and groups of correlated
features together (see methods) to calculate the effect of permuted OTUs
or group of OTUs on testing AUROC. The top 5 OTUs with the largest
negative impact on testing AUROC overlapped in tree-based models
{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
(OTU00367) abundances randomly, dropped the predictive performances the
most in all tree-based methods {[}Figure S3{]}. Decision tree, random
forest and XGBoost models' predictive performance dropped from 0.6 base
testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}.

We used permutation importance to interpret linear models as well to
highlight the differences between the two interpretation methods
{[}Figure S3{]}. The rankings of OTUs that discriminate between having
an SRN or not, changed between different interpretation methods.
L1-regularized SVM with linear kernel picked out some of the same OTUs
(OTU00822, OTU01239, OTU00609) as important in feature rankings
{[}Figure 3{]} and permutation importance methods {[}Figure S3{]} but
with different ranks. The difference was more pronounced in
L2-regularized models as they had mostly different OTUs picked out as
due to collinearity in microbial communities.

@@

As the complexity of a ML model and the number of tuned hyperparameter
settings increased {[}Table S1{]}, its training times increased as well
{[}Figure 5{]}. Linear models trained faster than non-linear models. L1
and L2 SVM with linear kernel and L2 logistic regression had training
times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
hours, (std ± 0.02), respectively. Whereas, a decision tree, SVM with
radial basis function kernel, random forest and xgboost had training
times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
(std ± 11.3) and 155.1 hours, (std ± 1), respectively {[}Figure 5{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. First, we
used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the differences between
cross-validation and test performance is low, this suggets the models
are not overfit and that they will perform similar with similar data. In
all seven models, the differences in cross-validation and testing AUROC
values did not exceed 0.021 which suggests that these models are
generalizable and can be used to test similar new data. Second, we
performed the initial 80\%-20\% random datasplit 100 times in our ML
pipeline. The randomization of the initial data-split to create a
held-out test set is a crucial step in the ML pipeline to develop robust
ML models and to report reliable performance metrics for a ML model.
Depending on the data-split, there is the chance of being overoptimistic
about the predictive performance of a model. In our study, we showed
that there was variability in AUROC values between different random
data-splits in each of the models we tested. Our results showed that the
testing AUROC values varied 0.23 on average, depending on the
data-split. Third, we used AUROC metric instead of accuracy in our study
to evaluate predictive performance of the ML models. AUROC is always
random at the value 0.5 and is a robust metric when a dataset is
imbalanced. We also performed a full grid search for hyperparameter
settings when building a ML model. It is not enough to use the default
hyperparameter settings when using previously developed ML packages in
programming languages such as R, Python and Matlab. In the example of
L1-regularized SVM with linear kernel (Figure S1), the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different
classification algorithms to show that we should use ML models based on
the goal of the study and our expectations of predictive performance,
interpretibility and computational burden. Microbiome studies use ML
models with the classification task to learn the training data to assign
labels, such as healthy or not, to new data (e.g.~new samples), but also
to learn which features are important enough to discriminate between
labels. If the goal of a study is not just to predict a health outcome
but also to learn the ecology behind a disease and to identify microbial
biomarkers of a disease, then interpretibility of a ML model is crucial.
In terms of predictive performance, random forest model had testing
AUROC values statistically significantly larger than others. However the
second best model was L2-regularized logistic regression following
random forest with a median AUROC difference of only 0.015. In terms of
interpretability, random forest was a more complex ML model and it could
only be explained using methods such as permutation importance. On the
other hand, L2-regularized logistic regression was easier to interpret
(i.e.~ranking absolute regression coefficients of the trained model).

Even with interpretable models, we need to be careful with how we treat
the information we get from the models. In this study we used two
methods to interpret our linear models; absolute feature weights of
trained models and permutation importance. We observed differences in
outcome between the two interpretation methods due to collinearity in
the dataset. Collinearity in a microbial dataset is when one OTU is
highly correlated with another OTU which causes regression coefficients
to not be unique. The regression coefficients of correlated OTUs are
influenced by one another which makes it difficult to interpret a ML
model. There are ways to make it easier to interpret models.; (1) use
different models for prediction and interpretation in which highly
correlated OTUs are removed, (2) get the highly ranked OTUs and check
what othe OTUs they are correlated with. The interpretation however
should not be a final answer but should be used to generate hypothesis
that are tested with followup experiments.

There are other criteria when choosing ML models such as the
computational burden of developing it, sample size and balance of the
dataset. In terms of computation burden, random forest model was slow to
train 83.2 hours whereas L2-regularized logistic regression trained in
12 minutes. The generalization performance of ML models depends on
sample size. The more complex the model, the more data it will need. The
dataset we used for our study had 490 samples, however microbiome
studies that have smaller sample sizes would benefit from using less
complex models such as L2-regularized logistic regression.

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (25). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(26--28).

@@

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For XGBoost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners.For hyperparameter selection, we started with a granular grid
search. Then we narrowed and fine-tuned the range of each
hyperparameter. The range of the grid depends on the ML task and ML
model. A full grid search needs to be performed to avoid variability in
testing performance. We can use hyper-band to help us with our
hyperparameter selection (29).

@@

\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
feature weights of (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel (C) L2 SVM with linear kernel were ranked from
highest rank 1 to 100 for each data-split. The feature ranks of the
highest ranked five OTUs based on their median ranks are shown here.
Similar OTUs had the largest impact on the predictive performance of L2
logistic regression and L2 SVM with linear kernel. Abbreviations: SVM,
support vector machine; OTU, Operational Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. The colors of the boxplots
stand for the unique OTUs that are shared among the different models;
pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for
OTU00110, green for OTU00361 and red for OTU00882. For all the
tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
the largest impact on predictive performance of the model.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for of each
data-split showed the differences in computational efficieny of the
seven models. The median training time in hours was the highest for
XGBoost and shortest for L1-regularized SVM with linear kernel. The
boxplot shows quartiles at the box ends and the statistical median as
the horizontal line in the box. The whiskers show the farthest points
that are not outliers. Outliers are data points that are not within 3/2
times the interquartile ranges. Abbreviations: AUROC, area under the
receiver operating characteristic curve; SVM, support vector machine;
XGBoost, extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2 logistic regression (B)
L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model.

@@

\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} (A) Decision tree (B)
Random forest (C) SVM with radial basis kernel (D) XGoost mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model. \newpage
\includegraphics{Figure_S4.png} \textbf{Figure S3. Interpretation of the
linear ML models with permutation importance.} (A) L1-regularized SVM
with linear kernel (B) L2-regularized SVM with linear kernel and (C)
L2-regularized logistic regression were interpreted using permutation
importance using held-out test set. The gray rectangle and the dashed
line show the IQR range and median of the base testing AUROC without any
permutation performed. Abbreviations: SVM, support vector machine; OTU,
Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational
Taxonomic Unit. \includegraphics{TableS1.pdf} \textbf{Table S1. ML model
complexity and hyperparameter range.} \newpage

@@

\hypertarget{ref-sze_leveraging_2018}{}
25. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
26. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-westcott_opticlust_2017}{}
27. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

>>>>

Machine learning (ML) modeling of the human microbiome has the potential
to identify the microbial biomarkers and aid in diagnosis of many
chronic diseases such as inflammatory bowel disease, diabetes and
colorectal cancer. Progress has been made towards developing ML models
that predict health outcomes from bacterial abundances, but rigourous ML
models are scarce due to the flawed methods that call the validity of
developed ML models into question. Furthermore, the use of black box ML
models has hindered the validation of microbial biomarkers. To overcome
these challenges, we benchmarked seven different ML models that use
fecal 16S rRNA sequences to predict the presence/absence of colorectal
cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To
show the effect of model selection, we assessed the predictive
performance, interpretability, and computational efficiency of the
following models: L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest, and extreme gradient boosting (XGBoost).
The random forest model was best at detecting CRC lesions with an AUROC
of 0.695 but it was slow to train (83.2 h) and hard to interpret.
Despite its simplicity, L2-regularized logistic regression followed
random forest in predictive performance with an AUROC of 0.680, and it
trained much faster (12 min). In this study, we showed that ML models
should be chosen based on expectations of predictive performance,
interpretability and available computational resources. Additionally, we
established standards for the development of modeling pipelines for
microbiome-associated ML models

@@

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. They learn from
existing data to predict the outcomes of new data and allow us to infer
on the reasons underlying that prediction. Therefore, researchers have
started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, the field's use of ML lacks clarity and consistency in
which methods are used and how these methods are implemented (12, 13).
More notably, we commonly see flawed ML practices such as (1) using ML
pipelines where there is no seperate held-out test dataset to evaluate
model performance or (2) reporting few or only the best outcomes of
cross-validation. Even when there are seperate testing sets to evaluate
model performance, there are large differences between cross-validation
and testing performances that indicate overfitting as well as large
confidence intervals for testing performances (4, 14--20).

Moreover, there is a lack of discussion on why a particular ML model is
utilized. Recently, there is a trend towards using more complex ML
models such as random forest, extreme gradient boosting and neural
networks without a discussion on if and how much model interpretibility
is necessary for the study (11, 21--23). Black box machine learning
models are not inherently interpretable and require posthoc explanations
to determine the feature importances in making a prediction. These
explanations can be misleading and at times unreliable when making
high-stake decisions about patient health (24). The models we develop
for healthcare, both to predict disease and to understand underlying
reasons behind that prediction, should be transparent and accountable
(25).

The lack of transparency on model selection and interpretation as well
as flawed modeling methods negatively impact model validity and
reproducibiliy. We need to strive toward better machine learning
practices by (1) implementing rigourous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of model accuracy, complexity, interpretibility
and computational efficiency.To showcase a rigorous ML pipeline and to
shed light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing several different ML models
using the same dataset and the same ML pipeline. We used a previously
published colorectal cancer (CRC) study (3) which had fecal 16S rRNA
gene sequences of 490 patients. The human-associated human microbiome is
hypothesized to directly contribute to the development of CRC and fecal
16S rRNA gene sequences have been used to detect it (1, 3, 4, 26). We
built seven ML models using fecal 16S rRNA gene sequences to predict
healthy patients versus patients with colorectal lesions that were
identified by colonoscopy as screen relevant neoplasias (SRN). The study
had 261 normal and 229 SRN samples. We established modeling pipelines
for L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels, a decision
tree, random forest and XGBoost. We used the area under the receiver
operating characteristic curve (AUROC) as a metric for determining the
predictibve performance of each model. The AUROC ranges from 1.0, where
the model perfectly distinguishes between cases and controls, to 0.50,
where the model's predictions are no different from random chance. The
median test AUROC varied from 0.601 to 0.695. Random forest had the
highest median AUROC for detecting SRN. Despite its simplicity, the
L2-regularized logistic regression was second best in predictive
performance. In terms of computational efficiency, L1 SVM with linear
kernel trained the fastest (0.202 hours, std ± 0.028), while XGBoost
took the longest (155.104 hours, std ± 0.959). We also found that
depending on how the data is split to create a held-out test set, the
AUROC values of a ML model can vary up to 0.295 which highlighted the
importance of performing many randomized modeling runs. This study
established standards for microbiome-associated ML models and
underscored the importance of model selection.

@@

We explained the feature importances in non-linear models; SVM with
radial basis kernel, decision tree, random forest and XGBoost, using a
metod called permutation importance on the held-out test data.
Permutation importance analysis is where we randomly permute
non-correlated features individually and groups of highly correlated
features together and calculate how much the predictive performance of
the model (i.e AUROC values) decrease if those specific OTUs or group of
OTUs are permuted randomly. We ranked the OTUs based on how much they
decreased the test AUROC values if they were randomly permuted, with the
largest decrease ranking highest. The top 5 OTUs with the largest
negative impact on testing AUROC overlapped in tree-based models
{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
(OTU00367) abundances randomly, dropped the predictive performances the
most in all tree-based methods {[}Figure S3{]}. Decision tree, random
forest and XGBoost models' predictive performance dropped from 0.6 base
testing median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models as well {[}Figure
S3{]}. The importance rankings of OTUs that discriminate between having
an SRN or not, changed between different interpretation methods.
L1-regularized SVM with linear kernel picked out some of the same OTUs
(OTU00822, OTU01239, OTU00609) as important in making a prediction in
the two interpretation methods; feature rankings based on weights
{[}Figure 3{]} and permutation importance {[}Figure S3{]}. Similarly,
L2-regularized SVM and L2-regularized logistic regression picked out
some of the same OTUs in both interpretation models, OTU00659 and
OTU00012, respectively. However, for all the linear models, the rankings
of these features were different due to the collinearity in microbial
communities.

@@

As the complexity of a ML model {[}Table S1{]} and the number of tuned
hyperparameter settings increased {[}Figures S1-S2{]}, its training
times increased as well {[}Figure 5{]}. Linear models trained faster
than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic
regression had training times of 0.2 hours, (std ± 0.03), 0.2 hours,
(std ± 0.02), and 0.2 hours, (std ± 0.02), respectively. Whereas, a
decision tree, SVM with radial basis function kernel, random forest and
xgboost had training times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ±
8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1), respectively
{[}Figure 5{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. First, we
used a held-out test set to illustrate the difference between
cross-validation and testing AUROC values. When the difference between
cross-validation and test performance is low, this suggets the models
are not overfit and that they will perform similar with similar data. In
all seven models, the difference between median cross-validation and
testing AUROC values did not exceed 0.021 which suggests that these
models are generalizable and can be used to test similar new data.
Second, we performed the initial 80\%-20\% random datasplit 100 times in
our ML pipeline. The randomization of the initial data-split to create a
held-out test set is a crucial step in the ML pipeline to develop robust
ML models and to report reliable performance metrics for a ML model.
Depending on how the data is split, there is the chance of being
overoptimistic about the predictive performance of a model. In our
study, we showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average
between different data-splits. Third, we used the AUROC metric instead
of accuracy in our study to evaluate the predictive performance of the
ML models. AUROC is always random at the value 0.5 and is a robust
metric when a dataset is imbalanced. We also performed a full grid
search for hyperparameter settings when building a ML model. Default
hyperparameter settings in previously developed ML packages in R,
Python, and Matlab programming languages are inadequate for effective
application of classification algorithms and need to be optimized for
each new dataset used to generate a model. In the example of
L1-regularized SVM with linear kernel {[}Figure S1{]}, the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

In this study, we benchmarked seven ML models with different
classification algorithms to show that we should use ML models based on
the goal of the study and our expectations of predictive performance,
interpretibility and computational burden. Microbiome studies use ML
models with a classification task to learn the training data to assign
labels, such as healthy or not, to new data, but also to learn which
features are important to discriminate between labels. If the goal of a
study is not just to predict a health outcome but also to learn the
ecology behind a disease and to identify microbial biomarkers of a
disease, then the ML model has to be interpretable. In terms of
predictive performance, random forest model had testing AUROC values
statistically significantly larger than others. However the second best
model was L2-regularized logistic regression following random forest
with a median AUROC difference of only 0.015. In terms of
interpretability, random forest was a complex ML model and it could only
be explained using methods such as permutation importance. On the other
hand, L2-regularized logistic regression was easier to interpret
(i.e.~ranking absolute regression coefficients of the trained model).

Even with interpretable models such as L2-regularized logistic
regression, we need to be careful with how we treat the information we
get from the models. In this study we used two different methods to
interpret our linear models; ranking each OTU by (1) their absolute
weights in the trained models and (2) their impact on the predictive
performance based on permutation importance. We observed differences in
the OTU rankings between the two interpretation methods due to
collinearity in the dataset. Collinearity in a microbial dataset is when
one OTU is highly correlated with another OTU which causes feature
weights to not be unique. The feature weights of correlated OTUs are
influenced by one another which makes it difficult to interpret the ML
model. To avoid misinterpreting the models we can; (1) use ML models
that has highly correlated OTUs removed, (2) use the highly ranked OTUs
that are correlated with other OTUs to generate hypotheses about the
ecology of the disease and test them with followup experiments.

There are other criteria when choosing ML models such as the
computational burden of developing it and sample size. In terms of
computational burden, random forest model trained each data-split in
83.2 hours whereas L2-regularized logistic regression trained in 12
minutes. The generalization performance of ML models depends on sample
size. The more complex the model, the more data it will need. The
dataset we used for our study had 490 samples, however microbiome
studies that have smaller sample sizes would benefit from using less
complex models such as L2-regularized logistic regression.

This study highlights the need to make educated choices at every step of
developing a ML model with microbiome data. Model selection, setting up
a rigorous ML pipeline and choosing the right interpretation method
contribute to achieving the level of validity and accountability we want
from models built for patient health.

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (27). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(28--30).

@@

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For XGBoost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners.For hyperparameter selection, we started with a granular grid
search. Then we narrowed and fine-tuned the range of each
hyperparameter. The range of the grid depends on the ML task and ML
model. A full grid search needs to be performed to avoid variability in
testing performance. We can use hyper-band to help us with our
hyperparameter selection (31).

@@

\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
feature weights of (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel (C) L2 SVM with linear kernel were ranked from
highest rank 1 to 100 for each data-split. The feature ranks of the
highest ranked five OTUs based on their median ranks are shown here.
Similar OTUs had the largest impact on the predictive performance of L2
logistic regression and L2 SVM with linear kernel. Abbreviations: SVM,
support vector machine; OTU, Operational Taxonomic Unit. \newpage
\includegraphics{Figure_4.png}

\textbf{Figure 4. Explanation of the non-linear ML models.} (A) SVM with
radial basis kernel (B) decision tree (C) random forest (D) XGboost
feaure importances were explanied using permutation importance using
held-out test set. The gray rectangle and the dashed line show the IQR
range and median of the base testing AUROC without any permutation
performed. The colors of the boxplots stand for the unique OTUs that are
shared among the different models; pink for OTU0008, salmon for OTU0050,
yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for
OTU00882. For all the tree-based models, a \emph{Peptostreptococcus}
species (OTU00367) had the largest impact on predictive performance of
the model. Abbreviations: SVM, support vector machine; OTU, Operational
Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic
Unit. \newpage
\includegraphics{Figure_5.png}

\textbf{Figure 5. Computational efficiency of seven ML models.} The
training times for of each data-split showed the differences in
computational efficieny of the seven models. The median training time in
hours was the highest for XGBoost and shortest for L1-regularized SVM
with linear kernel. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: AUROC,
area under the receiver operating characteristic curve; SVM, support
vector machine; XGBoost, extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png}

\textbf{Figure S1. Hyperparameter setting performances for linear
models.} (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2
SVM with linear kernel mean cross-validation AUROC values when different
hyperparameters are used in training the model. The differences in AUROC
values when hyperparameters change show that hyperparameter tuning is a
crucial step in building a ML model.

@@

\includegraphics{Figure_S2.png}

\textbf{Figure S2. Hyperparameter setting performances for non-linear
models.} (A) Decision tree (B) Random forest (C) SVM with radial basis
kernel (D) XGoost mean cross-validation AUROC values when different
hyperparameters are used in training the model. The differences in AUROC
values when hyperparameters change show that hyperparameter tuning is a
crucial step in building a ML model. \newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

\textbf{Figure S3. Interpretation of the linear ML models with
permutation importance.} (A) L1-regularized SVM with linear kernel (B)
L2-regularized SVM with linear kernel and (C) L2-regularized logistic
regression were interpreted using permutation importance using held-out
test set. The gray rectangle and the dashed line show the IQR range and
median of the base testing AUROC without any permutation performed.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.

\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
hyperparameter range.} \newpage

@@

\hypertarget{ref-rudin_optimized_2018}{}
25. \textbf{Rudin C}, \textbf{Ustun B}. 2018. Optimized scoring systems:
Toward trust in machine learning for healthcare and criminal justice.
Interfaces \textbf{48}:449--466.
doi:\href{https://doi.org/10.1287/inte.2018.0957}{10.1287/inte.2018.0957}.

\hypertarget{ref-knights_human-associated_2011}{}
26. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
signatures: Examining their predictive value. Cell Host Microbe
\textbf{10}:292--296.
doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.

\hypertarget{ref-sze_leveraging_2018}{}
27. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
